A netdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm

### 2025-07-29 paper.qmd
Adding final edits

---
title: "A NetdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format: pdf
keep-tex: true
bibliography: bibliography.bib
bibliographystyle: apa
link-citations: true
author: 
  - name: George G. Vega Yon
    affiliation:
      - name: University of Utah
    email: george.vegayon@utah.edu
  - name: Aníbal L. Olivera M.
    affiliation:
      name: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation:
      name: University of Southern California
# include-in-header: header.tex
---

@@

This paper introduces diffusion network research and presents an R library, netdiffuseR, that facilitates the estimation of contagion and diffusion processes on networks. Diffusion of innovations theory explains how new ideas and practices spread within and between communities. The spreading process consists of person-to-person interactions in which adopters persuade (by whatever means) non-adopters to adopt. Estimating these effects requires network data and an indicator for when each person adopts. The analytic steps involve creating an event history version of the data set and regression estimation to determine if exposure, the adopters in one’s network, is associated with individual adoption. NetdiffuseR automates many of these data management and processing steps, thus facilitating the estimation of these models. The package also contains simple ways to compute various network exposure measures such as based on structural equivalence, indirect ties, homophilic ties, etc. In addition, netdiffuseR provides capabilities to conduct simulations of network diffusion processes varying network structure, seed conditions, threshold distributions, and influence mechanisms. The paper includes example code for both empirical and simulated network studies. We aim to facilitate the adoption of netdiffuseR to advance the growth and evolution of diffusion network research.

@@

$$
E_{it} = \frac{\sum_{i\neq j}W_{ij}a_{j(t-1)}}{\sum_{i\neq j}W_{i+}}
$$

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; $a$ is an adoption vector indicating which of $i$’s $j$ alters had adopted at time $t - 1$; and the denominator is the row sum, the number of outgoing ties. $E_{it}$ in (1) measures the proportion of a person’s ties at time $t$ who have adopted the behavior at the prior time period, which is a contagion measure. As described in @Valente2024, however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), homophily, Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. NetdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

@@

The netdiffuseR R package [@netdiffuser_1; @netdiffuser_2] was first released to CRAN on February 18, 2016. The package contains a collection of algorithms and methods first envisioned by Dr. Valente and released in his book [@valente_book_1999] as GAUSS code. Since its release in 2016, the package has been downloaded over 45 thousand times and has about 400 hundred downloads a month, with an active community growing.

@@

The `diffnet` object we just created has various inspection methods. The print method gives a glimpse of the data, indicating the number of time points, nodes, type of network, and adoption prevalence, among other things. The summary method provides information about the adoption dynamics; it shows the number and proportion of adopters at each time point, network density, and Moran’s I [@moran_1950] indicating the level of spatial correlation. Finally, the `plot_diffnet()` function visualizes the diffusion process which colors nodes according to their adoption status, non-adopter, new adopter, and continuing adopter. The following code-chunk shows how to call these functions:

@@

Since exposure can be computed for each time point, the `exposure()` function will return a matrix of size $n \times \text{}$ the number of time points in the data; in the previous example, we had a glimpse of the exposure for the first ten individuals at time point five.

@@

The function is a wrapper of the `glm()` function in R that allows estimating Generalized Linear Models. In our case, `diffreg()` estimates a logistic regression model that predicts adoption as a function of network exposure. Furthermore, as it can be seen in the output, the function automatically truncates the data up to the adoption time (like a survival analysis). Importantly, the function throws a warning message indicating that the `per` variable has not been included in the model, which can lead to bias estimates.

@@

```{r}
#| label: simulating-adoption-process-3
# Creating a graph based on watts-strogatz model
set.seed(121)
graph <- rgraph_ws(200, 10, p=.3)

# Run the simulation
diffnet_3 <- rdiffnet(
  t              = 10,
  seed.graph     = graph,
  seed.p.adopt   = 0.1,
  threshold.dist = runif(200, .3,.5)
)
```

@@

Starting with version 1.23.0, **netdiffuseR** supports simulating multi- and dis-adoption diffusion processes. To study a multi-adoption process, you can pass a `list` as the `seed.p.adopt` parameter. Here is a simple example using the Koren Family Planning network:

```{r}
#| label: simulation-multiadoption-process-4
set.seed(1231)

diffnet_4 <- rdiffnet(
 200, 10,
 seed.p.adopt = list(0.1, 0.15)
 )

@@

```{r}
#| label: simulation-multiadoption-process-6
diffnet_6 <- rdiffnet(
  200, 10,
  seed.p.adopt = list(0.1, 0.15),
  threshold.dist = list(
    runif(200, .3, .5),
    runif(200, .2, .4)
  ),
  seed.nodes = list("central", "random"),
  behavior   = list("tobacco", "alcohol")
)

@@

As we did in the previous section, we can give a specific network as input. Besides passing fixed networks as we did with the small-world example, the `rdiffnet` function also supports passing diffnet objects as input. When doing so, the function will use the graph of the diffnet object as the seed graph and will take the time argument as the number of timepoints included in the graph:

```{r}
#| label: simulation-multiadoption-process-7
diffnet <- rdiffnet(
  seed.graph     = kfam_diffnet,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(nvertices(kfam_diffnet), .3,.5)
  )

diffnet
```

@@

```{r}
#| label: simulation-multiadoption-process-7-plot
diffnets <- split_behaviors(diffnet)
op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets[[1]], main = "Behavior 1")
plot_adopters(diffnets[[2]], main = "Behavior 2")
par(op)
```

@@

The disadoption feature included in **netdiffuseR** version 1.24.0 opened a new way of studying network diffusion processes. Considering the disadoption of an innovation or behavior is essential for studying significant aspects of competition between products or beliefs [@lehmann_disadoption_2017]. The `rdiffnet()` function includes the `disadopt` parameter to add a disadoption function, facilitating such analyses and enabling the testing of some theoretical models for disadoption [@alipour_enough_2024].

@@

2. It must return a list with as many elements as behaviors there are: It should list which nodes are dropping behaviors. For instance, if only node `1` (one) is dropping behavior one of three, it should return `list(1, integer(), integer())`.

3. If there are no nodes to disadopt or adopt, the function must return an empty vector (`integer()`).

@@

```{r}
#| label: disadoption-example-1
diffnet_7  <- rdiffnet(
  seed.graph   = graph,
  t            = 10,
  disadopt     = random_dis,
  seed.p.adopt = 0.1
  )

diffnet_7
```

@@

1. We are identifying individuals adopting more than one behavior at time `t` using the code `apply(cumadopt[, time,], 1, sum) > 1`. In a two behavior model, this will return a vector with values 0 (no adoption), 1 (only one behavior adopted), or 2 (both behaviors adopted).

@@

3. If there are no double adopters, the function returns an empty list.

@@

```{r}
#| label: disadoption-example-2
set.seed(331)
diffnet_8 <- rdiffnet(
  200, 10,
  disadopt = one_only,
  seed.p.adopt = list(0.1, 0.1)
)
diffnet_8
```

@@

```{r}
#| label: checking-if-if-worked
#
toas <- toa_mat(diffnet_8)

@@

This paper has provided an introduction to the NetdiffuseR R package designed to facilitate estimating contagion and diffusion processes on networks. The main goal was to provide example code and tutorial so others may use the package to explore diffusion network models either with the existing classic datasets that come with the package or using their own data. The main tutorial sections included sample code for estimating network exposure, the workhorse variable in network research. Simulation features of netdiffuseR were illustrated which provide tools for theoretical testing of diffusion processes. We also provided examples of extensions to exposure that provide a wide array of different weighting techniques, such as structural equivalence, homophilic ties, and so on.

@@

We hope that by providing these illustrative examples, others will be able to use NetdiffuseR in their work. Diffusion processes are ubiquitous in science and everyday life. The diffusion paradigm was first established by Ryan and Gross in 1943 (Valente & Rogers, 1995) but it continues to grow and evolve, and our hope is that this publication will further encourage both growth and evolution.

>>>>

---
title: "A netdiffuseR Tutorial: Estimating Network Influences on Behavior within the Diffusion Paradigm"
format: pdf
keep-tex: true
bibliography: bibliography.bib
bibliographystyle: apa
link-citations: true
author: 
  - name: George G. Vega Yon
    affiliation:
      - name: University of Utah
    email: george.vegayon@utah.edu
  - name: Aníbal L. Olivera M.
    affiliation:
      name: Universidad del Desarrollo
  - name: Thomas W. Valente
    affiliation:
      name: University of Southern California
# include-in-header: header.tex
---

@@

This paper introduces diffusion network research and presents an R library, netdiffuseR, that facilitates the estimation of contagion and diffusion processes on networks. Diffusion of innovations theory explains how new ideas and practices spread within and between communities. The spreading process consists of person-to-person interactions in which adopters persuade (by whatever means) non-adopters to adopt. Estimating these effects requires network data and an indicator for when each person adopts. The analytic steps involve creating an event history version of the data set and regression estimation to determine if exposure, the adopters in one’s network, is associated with individual adoption. netdiffuseR automates many of these data management and processing steps, thus facilitating the estimation of these models. The package also contains simple ways to compute various network exposure measures such as based on structural equivalence, indirect ties, homophilic ties, etc. In addition, netdiffuseR provides capabilities to conduct simulations of network diffusion processes varying network structure, seed conditions, threshold distributions, and influence mechanisms. The paper includes example code for both empirical and simulated network studies. We aim to facilitate the adoption of netdiffuseR to advance the growth and evolution of diffusion network research.

@@

$$
E_{it} = \frac{\sum_{i\neq j}W_{ij}a_{j(t-1)}}{\sum_{i\neq j}W_{ij}}
$$

Where $W_{ij}$ is a weight matrix indicating ties between individuals, most commonly a binary adjacency matrix such as who is friends with whom but can be any relation; $a$ is an adoption vector indicating which of $i$’s $j$ alters had adopted at time $t - 1$; and the denominator is the row sum, the number of outgoing ties. $E_{it}$ in (1) measures the proportion of a person’s ties at time $t$ who have adopted the behavior at the prior time period, which is a contagion measure. As described in @Valente2024, however, there can be many extensions of the exposure calculation by estimating influence of ties based on: a count (removing the denominator), indirect connectivity, structural equivalence, strength (weak vs. strong), homophily, Simmelian relations (alter-to-alter), alter network properties such as centrality, etc. netdiffuseR enables users to calculate these alternative exposures with a parameter toggle in the exposure function.

@@

The netdiffuseR R package [@netdiffuser_1; @netdiffuser_2] was first released to CRAN on February 18, 2016. The package contains a collection of algorithms and methods first envisioned by Dr. Valente and released in his book [@valente_book_1999] as GAUSS code. Since its release in 2016, the package has been downloaded over 45 thousand times and has about 400 downloads a month, with an active community growing.

@@

The `diffnet` object we just created has various inspection methods. The print method gives a glimpse of the data, indicating the number of time points, nodes, type of network, and adoption prevalence, among other things. The `summary` method provides information about the adoption dynamics; it shows the number and proportion of adopters at each time point, network density, and Moran’s I [@moran_1950] indicating the level of spatial correlation. Finally, the `plot_diffnet()` function visualizes the diffusion process which colors nodes according to their adoption status, non-adopter, new adopter, and continuing adopter. The following code-chunk shows how to call these functions:

@@

Since exposure can be computed for each time point, the `exposure()` function will return a matrix of size $n \times \text{}$ the number of time points in the data; in the previous example, we had a glimpse of the exposure for the first six individuals at time point five.

@@

The function is a wrapper of the `glm()` function in R that allows estimating Generalized Linear Models. In our case, `diffreg()` estimates a logistic regression model that predicts adoption as a function of network exposure. Furthermore, as it can be seen in the output, the function automatically truncates the data up to the adoption time (like a survival analysis). Importantly, the function throws a warning message indicating that the period variable, `per`, has not been included in the model, which can lead to bias estimates.

@@

```{r}
#| label: simulating-adoption-process-2
# Creating a graph based on watts-strogatz model
set.seed(121)
graph <- rgraph_ws(200, 10, p=.3)

# Run the simulation
diffnet_2 <- rdiffnet(
  t              = 10,
  seed.graph     = graph,
  seed.p.adopt   = 0.1,
  threshold.dist = runif(200, .3,.5)
)
```

@@

Starting with version 1.23.0, **netdiffuseR** supports simulating multi- and dis-adoption diffusion processes. To study a multi-adoption process, you can pass a `list` as the `seed.p.adopt` parameter. Here is a simple example:

```{r}
#| label: simulation-multiadoption-process-3
set.seed(1231)

diffnet_3 <- rdiffnet(
 200, 10,
 seed.p.adopt = list(0.1, 0.15)
 )

diffnet_3
```

@@

```{r}
#| label: simulation-multiadoption-process-4
diffnet_4 <- rdiffnet(
  200, 10,
  seed.p.adopt = list(0.1, 0.15),
  threshold.dist = list(
    runif(200, .3, .5),
    runif(200, .2, .4)
  ),
  seed.nodes = list("central", "random"),
  behavior   = list("tobacco", "alcohol")
)

@@

As we did in the previous section, we can give a specific network as input. Besides passing fixed networks as we did with the small-world example, the `rdiffnet()` function also supports passing `diffnet` objects as input. When doing so, the function will use the graph of the `diffnet` object as the seed graph and will take the time argument as the number of timepoints included in the graph. Here is a simple example using the Koren Family Planning network:

```{r}
#| label: simulation-multiadoption-process-5
diffnet_5 <- rdiffnet(
  seed.graph     = kfam_diffnet,
  seed.p.adopt   = list(0.1, 0.15),
  threshold.dist = runif(nvertices(kfam_diffnet), .3,.5)
  )

diffnet_5
```

@@

```{r}
#| label: simulation-multiadoption-process-5-plot
diffnets_5 <- split_behaviors(diffnet_5)
op <- par(mfrow=c(1,2), cex = .8)
plot_adopters(diffnets_5[[1]], main = "Behavior 1")
plot_adopters(diffnets_5[[2]], main = "Behavior 2")
par(op)
```

@@

The disadoption feature included in **netdiffuseR** version 1.23.0 opened a new way of studying network diffusion processes. Considering the disadoption of an innovation or behavior is essential for studying significant aspects of competition between products or beliefs [@lehmann_disadoption_2017]. The `rdiffnet()` function includes the `disadopt` parameter to add a disadoption function, facilitating such analyses and enabling the testing of some theoretical models for disadoption [@alipour_enough_2024].

@@

2. It must return a list with as many vectors as behaviors there are: It should list which nodes are dropping behaviors. For instance, if only node `5` (five) is dropping behavior one of three, it should return `list(c(5), integer(), integer())`.

3. If there are no nodes to disadopt or adopt none of $n$ behaviors, the function must return a list with empty vectors: `replicate(n, integer())`.

@@

```{r}
#| label: disadoption-example-1
diffnet_6  <- rdiffnet(
  seed.graph   = graph,
  t            = 10,
  disadopt     = random_dis,
  seed.p.adopt = 0.1
  )

@@

1. We are identifying the number of behaviors adopted by each individual at time `t` using the code `apply(cumadopt[, time,], 1, sum)`. In a two behavior model, this will return a vector with values 0 (no adoption), 1 (only one behavior adopted), or 2 (both behaviors adopted).

@@

3. If there are no double adopters, the function returns a list with empty vectors.

@@

```{r}
#| label: disadoption-example-2
set.seed(331)
diffnet_7 <- rdiffnet(
  200, 10,
  disadopt = one_only,
  seed.p.adopt = list(0.1, 0.1)
)
diffnet_7
```

@@

```{r}
#| label: checking-if-if-worked
#
toas <- toa_mat(diffnet_7)

@@

This paper has provided an introduction to the netdiffuseR R package designed to facilitate estimating contagion and diffusion processes on networks. The main goal was to provide example code and tutorial so others may use the package to explore diffusion network models either with the existing classic datasets that come with the package or using their own data. The main tutorial sections included sample code for estimating network exposure, the workhorse variable in network research. Simulation features of netdiffuseR were illustrated which provide tools for theoretical testing of diffusion processes. We also provided examples of extensions to exposure that provide a wide array of different weighting techniques, such as structural equivalence, homophilic ties, and so on.

@@

We hope that by providing these illustrative examples, others will be able to use netdiffuseR in their work. Diffusion processes are ubiquitous in science and everyday life. The diffusion paradigm was first established by Ryan and Gross in 1943 [@valente_origins_1995] but it continues to grow and evolve, and our hope is that this publication will further encourage both growth and evolution.
